---
title: "Meshes are needed <br>for geospatial work"
output: ioslides_presentation
editor_options: 
  chunk_output_type: console
---



<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
  .col3 {
    columns: 3 100px;
    -webkit-columns: 3 100px;
    -moz-columns: 3 100px;
  }
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(rgl)
library(dplyr)
library(raster)
library(silicate)
```

## Questions

* What are meshes and topology?
* Relationship of meshes to raster/vector data?

## What is a mesh?

Key ideas!  

* **Topology**: the *shape* of things and their relationships.  
* **Geometry**: the *where* of things. 
* **Indexing**: the *link* between topology and geometry. 

<br><br><br><br><br><br><br><br>
*(indexing of data-structures*, not database-indexing)

## Topology is not geometry

Topology emerges as an *index* into geometry. 

```{r topology0,echo=TRUE}
## this is pure 3-space geometry, no connections
(geometry <- cbind(x = c(0, 0.5, 1), y = c(0, 0.5, 1), z = c(0, 0, 0.8)))

# line of 2-segments has 1-dimensional topology (2-index) 
(topology1 <- cbind(.v0 = c(1, 2), .v1 = c(2, 3)))
```

## Topology is not geometry 

```{r topology1,echo=TRUE}
## this is pure 3-space geometry, no connections
geometry

# triangle has 2-dimensional topology (3-index)
(topology2 <- cbind(.v0 = 1, .v1 = 2, .v2 = 3))
```

## Shared geometry

```{r topology2,echo=FALSE}
#Topology vs. geometry

library(rgl)
rgl.clear()
mfrow3d(1, 2, sharedMouse = TRUE)
library(rgl)
lines3d(geometry[t(topology1), ], lwd = 3, col = "firebrick")
material3d(col = "black")
axis3d("x")
axis3d("y")
axis3d("z")
title3d(xlab = "x", ylab = "y", zlab = "z")
quads3d(cbind(c(0, 1, 1, 0), c(0, 0, 1, 1), c(0, 0, 0, 0) - 0.01), col="gray")

next3d()
triangles3d(geometry[t(topology2), ], col = "firebrick")
material3d(col = "black")
axis3d("x")
axis3d("y")
axis3d("z")
title3d(xlab = "x", ylab = "y", zlab = "z")
quads3d(cbind(c(0, 1, 1, 0), c(0, 0, 1, 1), c(0, 0, 0, 0)), col="gray")
rglwidget()
```

## What is wrong with spatial? 

*Traditional spatial data tends to confuse topology and geometry.*  

*Polygons are topologically equivalent to lines.* 

*How to triangulate a polygon?*   


## What is wrong with spatial? 

*Traditional spatial data tends to confuse topology and geometry.*  

*Polygons are topologically equivalent to lines.* 

*How to triangulate a polygon?*           (YIKES)

*Raster and vector aren't so different*. 

## Motivations 

* Lossless raster reprojection
* Topology fixes and neighbour classification
* Tracks and point clouds
* 3D and 4D visualization, immersive experiences

I want to jump the chasm from geospatial to computer graphics and VR. 


## What is a raster? 

What is a raster?

A layer of neighbouring rectangles?   A regular grid of points? 

```{r what-raster, echo=FALSE}
library(raster)
plot_edges <- function(x, main = "") {
  sc <- silicate::SC(spex::polygonize(x))
  e <- silicate::sc_edge(sc)
  v <- silicate::sc_vertex(sc)
  x0 <- e %>% dplyr::inner_join(v, c(".vx0" = "vertex_"))
  x1 <- e %>% dplyr::inner_join(v, c(".vx1" = "vertex_"))
  plot(rbind(x0, x1)[c("x_", "y_")], asp = 1, type = "n", 
       axes = FALSE, xlab = "", ylab = "", main = main)
  graphics::segments(x0$x_, x0$y_, x1$x_, x1$y_, lty = 2)
}
## What is a raster? 
m <- matrix(c(seq(0, 0.5, length = 5), 
            seq(0.375, 0, length = 4)), 3)
colpal <- function(n = 26, drop = 5) sf::sf.colors(n)[-seq_len(drop)]
plot_values <- function(x) {
  plot(x, axes = FALSE, main = "value", box = FALSE, col = colpal()); 
  text(coordinates(x), label = values(x))
  plot(extent(x), add = TRUE)
}
plot_cells <- function(x) {
  plot(x, axes = FALSE, main = "cell", legend = FALSE, box = FALSE, col = colpal()); 
  plot(extent(x), add = TRUE)
  text(coordinates(x), label = sprintf("[%i]", seq_len(ncell(x))), cex = 0.8)
}
r <- setExtent(raster(m), extent(0, ncol(m), 0, nrow(m)))
op <- par(mfcol = c(1, 2))
plot_values(raster(m))
plot_edges(r, main = "points")
points(coordinates(r), col = colpal(10, drop = 1)[scales::rescale(values(r), c(1, 9))], pch = 19, cex = 1.5)

par(op)
```




## What is a polygon? 

A series of grouped *paths*?

```{r, echo=FALSE}
library(silicate)
mmcol <- c( "#0000B3FF", "#FFF50AFF") #viridis::viridis(2)
mm <- sf::as_Spatial(minimal_mesh)
par(mfcol = c(1, 2))
plot(mm, main = "two polygons", col = mmcol)
plot(mm, border = "transparent", main = "three paths")
sc <- silicate::SC0(mm)
o <- tidyr::unnest(sc$object[2:1, ]) ## flip so purple on top
x0 <- sc$vertex[o$.vx0, ]
x1 <- sc$vertex[o$.vx1, ]
for (i in seq_len(nrow(x0))) {  ## sigh
  arrows(x0$x_[i], x0$y_[i], x1$x_[i], x1$y_[i], col = mmcol[o$a[i]], lwd = 2, length = 0.1, angle = c(30, 15)[o$a[i]])
}
```

## What's the fill? 

In 3D, need to fill our polygons properly *as data*. 

```{r, echo=FALSE}
tri <- sfdct::ct_triangulate(minimal_mesh, a = 0.008, D = TRUE)
geom <- tri$geom
tri <- sf::st_set_geometry(tri, NULL)

tri <- tri[rep(1:2, lengths(geom)), , drop = FALSE]

tri[["geom"]] <- sf::st_as_sfc(lapply(unlist(geom, recursive = FALSE), sf::st_polygon))
tri <- sf::st_as_sf(tibble::as_tibble(tri))


## repeat previous plot
op <- par(mfcol = c(1, 2))
plot(mm, main = "two polygons", col = mmcol)
plot(mm, border = "transparent", main = "what goes here?")
sc <- silicate::SC0(mm)
o <- tidyr::unnest(sc$object[2:1, ]) ## flip so purple on top
x0 <- sc$vertex[o$.vx0, ]
x1 <- sc$vertex[o$.vx1, ]
for (i in seq_len(nrow(x0))) {  ## sigh
  arrows(x0$x_[i], x0$y_[i], x1$x_[i], x1$y_[i], col = mmcol[o$a[i]], lwd = 2, length = 0.1, angle = c(30, 15)[o$a[i]])
}
#coords <- sf::st_coordinates(sf::st_centroid(tri))
coords <- coordinates(as(tri, "Spatial"))
text(coords, label = "?", cex = 0.7, col = mmcol[(tri$a == 2) + 1])
#text(0.35, 0.35, "just\n a hole!", cex = 0.5)
par(op)
```

The *fill* we see in traditional 2D graphics is a **trick!!**. 


## Triangles

To fill our polygon we need triangles. 

```{r}
#tri <- sf::st_cast(sfdct::ct_triangulate(minimal_mesh, a = 0.01, D = TRUE))
## repeat previous plot
op <- par(mar = par("mar") * 2)
plot(tri, main = "two polygons", col = rep(mmcol, lengths(geom)))
par(op)
```

A raster and vector mesh use the same concepts. 

## When things go wrong

```{r broken,include=FALSE}
library(silicate)
borked <- minimal_mesh
borked$geom[[2]][[1]] <- list(structure(c(0.69, 0.8, 1.1, 1.23, 1.23, 1.23, 0.69, 0, 0.6, 0.63, 0.3, 0.1, 0.3,
0), .Dim = c(7L, 2L), .Dimnames = list(NULL, c("x", "y"))))
#plot(borked)
```

`st_is_valid(x[2, ]) == FALSE`

```{r sf-valid,echo-FALSE,include=TRUE}
plot(as(borked, "Spatial"), col = mmcol)
#sf::st_is_valid(borked)

```

## R package silicate, topology in tables
<div class="col2">

````{r silicate,echo=FALSE,include=TRUE,out.width="100%"}
library(silicate)
mesh <- SC0(borked)
vertex <- mesh$vertex
vertex$v <- 1:nrow(vertex)
vertex <- vertex[sort(unlist(mesh$object$topology_[[2]])), ]
plot(vertex$x_, vertex$y_, asp = 1, pch = ".")
text(vertex$x_, vertex$y_, lab = vertex$v, pos = 2, adj = 1)
idx<- mesh$object$topology_[[2]]
segments(mesh$vertex$x_[idx[[1]]], mesh$vertex$y_[idx[[1]]], 
         mesh$vertex$x_[idx[[2]]], mesh$vertex$y_[idx[[2]]])

```
```{r edges}
mesh$object$topology_[[2]]
```


## Live code


[http://rpubs.com/cyclemumner/geomesh-r](http://rpubs.com/cyclemumner/geomesh-r)



## Thanks!

Yakkin: @mdsumner

Github: [github.com/mdsumner](https://github.com/mdsumner)

Dreams: [github.com/hypertidy](https://github.com/hypertidy)

Australian Antarctic Division  [antarctica.gov.au](https://www.antarctica.gov.au)

Antarctic Climate and Ecosystems CRC  [acecrc.org.au](http://www.acecrc.org.au)

[rOpenSci.org](https://www.rOpenSci.org)

